Fish do not need Visas: the transboundary nature of the world’s exploited marine fish species

### 2019-12-17 Manuscript_Draft.Rmd
Responded to William comments

---
title: "Fish do not need Visas: the transboundary nature of world fished species"
author: "Authors: Juliano Palacios-Abrantes^1^, Gabriel Reygondeau^1,2^, Colette Wabnitz^3^ and William W.L. Cheung^1^"
date: "Affiliations: ^1^Institute for the Oceans and Fisheries, University of British Columbia, Vancouver, Canada, ^2^Yale University, New Haven, Connecticut, USA, ^3^Stockholm Resilience Centre, Stockholm University, Stockholm, Sweden"
subtitle: "DRAFT DO NOT SHARE"
# subtitle: "Fish do not need Visas (16/40 character maximum)"
output:
  word_document:
    # reference_docx: ./Manuscript_Instructions/science_template.docx
  pdf_document:
    fig_caption: yes
geometry: margin=1in
editor_options:
  chunk_output_type: inline
# bibliography: ./References/FishForVisa_Ref.bib
# csl: ./References/science.csl
---

@@

Species distribution around the world is not a random process. The area in space and time that a species population occupies is a response to the availability of resources, a series of biotic and abiotic factors, as well as their influence on the evolutionary processes [@Hunch]. Marine resource management is predicated on knowledge of species distributions as human-made spatial boundaries are inherent to existing management systems (e.g., marine protected areas) [@Song:2017iua]. However, incorporating the concept of species distribution into fisheries management, or policy, is a relatively recent construct [@Alexa]. The delineation of Economic Exclusive Zones (EEZs) in the early 80s, for instance, virtually established boundaries across the distribution of many species, creating shared species between nations, and adding another level of complexity to fisheries management. Forty years after their formal adoption, the question of how many shared species exist in the world based on their distribution is still unanswered. Fishing quotas set by different nations for a shared fishery resource have led to conflict [@Spijkers:2017ij]. Climate change, which is shifting species distributions, is likely to exacerbate this type of conflict [@Pinsky:2018cb] and presents further challenges for fisheries management (@Mills). Thus, answering the questions of how many shared stocks exist in the world, and where these are located, is critical to developing effective and resilient management plans for the continued and future sustainable use and conservation of marine resources.

From 1973 to 1982, members of the United Nations held a series of meetings to discuss regulations regarding the high seas, a region of international common property, at that time, consisting of waters from 12 miles from shore [@McRaeandMunro1989]. The establishment of the UN Convention on the Law of the Sea (UNCLOS) allowed coastal states to claim jurisdiction over the exploration and exploitation of marine resources over 200 nautical miles off their coasts [@UN:1982]. While intended to improve fisheries management by granting property rights over shared resources, this arbitrary delimitation of management areas was not informed by bio-geography, essentially ignoring species distributions and creating what we know today as shared fish stocks [@Gulland, @Munro2002]. Fish stocks, are comonly refered in fisheries management as a particular population that is more or less isolated from other stocks of the same species and hence self-sustaining [@FAO:1999tk]. The Food and Agricultural Organization (FAO) recognizes three types of shared stocks: (*i*) transboundary stocks, shared by 2 or more neighboring coastal nations; (*ii*) straddling stocks, occurring in two or more adjacent national jurisdictions and the high seas; and (*iii*) highly migratory stocks, found in the EEZs of coastal nations that are not necessarily adjoining, and the high seas. The inception of shared stocks called for the establishment of new fisheries management methods. Managers and scientists adopted a game theory approach – or means to analyze strategic interactions among decision-makers – which demonstrates that collaboration is most likely to result in the best overall outcome for nations sharing a common resource [@Bailey:2010ga,@Eide:2013dh]. Lack of collaboration in shared stocks may threaten stock sustainability [@Clark:1980js; @Nguyen:2018dk], reduce the profitability potential of the fishery [@Merino:2007jz], and lead to conflict between coastal nations [@Spijkers:2017ij], although cooperation and non-cooperation can be more complex [@Jensen:2015cf]. Moreover, under climate change, species distributions will shift [@Cheung2016], resulting in new shared stocks [@Pinsky:2018cb] and highlighting the key role cooperative fisheries management can play in maintaining stocks and profitability [@Nguyen:2018dk] as well as hopefully reducing fisheries conflict [@Spijkers:2019cz].

There is a gap in assessing the nature and number of transboundary species in the world based on their spatial distribution. Caddy [@Caddy:1997ue] in 1997 estimated that there could be up to 1,500 transboundary fish species in the world. However, such estimation lacked a proper assessment due to limited information on the distribution of species at the time [@Caddy:1997ue]. More recently, Teh et al. [@Teh:2015gd] estimated that 206 species can be considered shared, accounting for a total catch of 34.2 million t and a global landed value of USD 30.7 trillion. While this represents a good starting point, the analysis was based on a literature review of shared species and did not include a mechanistic way of determining the transboundary nature of the species [@Teh:2015gd]. Thus, despite recent research highlighting the interconnection of marine species [@Ramesh:2019va], the total number of transboundary species is yet to be identified at a global scale. Yet, such an understanding is critical to effectively managing fisheries, as needs differ substantially between discreet and shared species [@Mills] and climate change will yield new challenges associated with new shared fisheries [@Pinsky:2018cb].

 <!-- **Methods summary** (250/250 words) -->

In this study, we overlay the known distribution of **968** commercially valuable marine fish species and the Exclusive Economic Zones of 175 coastal countries to determine the total number of transboundary species and their contribution to global catch and landed value. We adopt the UN definition of transboundary stocks (shared by neighboring EEZs), but carry out the analysis at the species level, rather than stocks (populations within a species), due to lack of stock-specific spatial and ecological information on all of the fished species [@Teh:2015gd].

We combined multiple species distribution data-sources and developed a series of three criteria to determine if a species is transboundary. Current species distributions were obtained from the largest observational database of the world’s marine species [@REFNEreus] and two structurally distinct species distribution models [@REFGabs; @Zeler]. The first criteria to determine the transboundary nature of a species considered the number of data-sources agreeing that a species was present in a single grid-cell. Second, we used spatially explicit catch data [@Zeller] to confirm the occurrence of a species in a given grid cell. The species was droped if it was present in any specific grid cell, but no catch was record was found. The third criteria considered the spatial distribution of species between neighboring EEZs. In addition to determining the number of transboundary stocks, we also estimated the contribution of each to fishing revenues and landings as well as categorized them according to catch trends [@Kleisner:2013gh]. Our study provides a first empirical approximation of the number of shared commercial species globally, significantly contributing to our understanding of the true transboundary nature of world fisheries and hopefully helping inform the effective management of these resources.

<!-- **Results and discussion** (~500/1000 words) -->

We estimate a total of 633 transboundary species in the world, more than double the number previously estimated [@Teh:2015gd] (Fig. 1A). We also find that fisheries on these species contribute substantially to both landings and revenue. Between 2005 and 2014, fleets targeting shared stocks within EEZs landed an average of 69 million tonnes, representing around 64% of the global catch (within EEZs), and generating USD 79.7 million in fishing revenue, equivalent to 49.8% of total revenue. These values are higher than previously estimated [@Teh]. Transboundary species are particularly important in terms of catch and revenue in regions of the world like Eastern Asia and Latin America and the Caribbean (Fig 1B). Particular in Northern America (Bermuda, Canada, Greenland, and the US), a moderate amount of catch (3.4 thousand tons) are worth 11.7 billion USD, highlighting the importance of transboundary species to the economy of the region. This is the case of the US where 116 species are responsible for up to 12.20 million USD and Peru where 28 species are responsible for USD 5.2 million (Fig 1A).

@@

We found transboundary fisheries to be of substantial economic importance in regions with historic fisheries-related conflicts. The Asian region with a total number of **310** transboundary species contributing  **20'.4 million tons** to the region's total landings and **19.3 billion USD** in revenue has experienced 43.0% of all documented international fisheries conflicts since 2000 [@Spijkers:2019cz]. The European Union, Norway, Iceland, and the Faroe Islands have been at odds over the size and relative allocation of total allowable catches (TACs) for mackerel since 2007 [@Spijkers:2017ij], a species that is transboundary and contributes **xx** to landings and **xx** in revenue to the countries involved in the dispute. The Mediterranean, a region that has experienced conflicts since the 1980s [@Kliot:1989kv], is home to a large number of transboundary stocks contributing over 6 million tones worth over **8 billion USD** to countries along the Sea's rim. The identification of current transboundary species is a first step towards establishing joint management plans for marine resources [@Link:2010ei], key not only to sustain fisheries over the long term, but also to avoid international conflict.

The two regions of the world that share the most species (the North East and Central Atlantic) have developed quite different approaches to managing shared stocks. In the North East Atlantic, fisheries are managed by the European Union and quotas allocated by species and regulatory area, with some of them (e.g., Iceland and Norway) characterized as having good fisheries management [@Melnychuk:2017gh]. On the other hand, fisheries along the west coast of Africa could benefit from improvements [@Melnychuk:2017gh]. Ironically, this region has long been exploited by foreign fleets, including European, sometimes illegally [@Belhabib:2015bl], with questionable fishing rights systems [@Mills:2004dj], and unfair fisheries agreements [@Kaczynski & Fluharty 2002]. Thus, having joint management plans for shared stocks between neighbors will not be enough to achieve sustainable goal 14, Life Below Water [@SDG]. Countries need to work together towards a more equitable system of global fishing rights.

@@

We present a novel method to determine whether or not a species can be considered transboundary. The method is based on a series of criteria to be met and the overlay of species distributions and the spatial boundaries of the world’s EEZs. For the current analysis we work at the species level and adopt the Food and consider transboundary species as those that occur within the Exclusive Economic Zone (EEZ) of neighboring countries [@Munro:2002uf; @FAO]. All data was scaled to 0.5 degree latitude x 0.5 degree longitude grid cells (*n* = ~180,000 grid cells).

@@

We use the occurrence database, the ENM-Nereus model, and SDM-SAU model to determine the presence of each species within each of the 0.5 x 0.5 grid cells. We computed a *Species Index* by dividing the number of datasets that confirm presence of a species in a grid cell over the total amount of datasets (*n* = 3). The index is unit-less and represents a scale from 0.3 to 1 where 0.3 means only one dataset reports presence of the species in a grid cell and 1 means agreement between all datasets. We selected only those cases where *Species Index* = 1 to get a more conservative estimate of transboundary species (**S1_Map_per_dataset**).

@@

>>>>

---
title: "Fish do not need Visas: the transboundary nature of the world’s exploited marine fish species"
author: "Authors: Juliano Palacios-Abrantes^1^, Gabriel Reygondeau^1,2^, Colette Wabnitz^1,3^ and William W.L. Cheung^1^"
date: "Affiliations: ^1^Institute for the Oceans and Fisheries, University of British Columbia, Vancouver, Canada, ^2^Yale University, New Haven, Connecticut, USA, ^3^Stockholm Resilience Centre, Stockholm University, Stockholm, Sweden"
subtitle: "DRAFT DO NOT SHARE"
# subtitle: "Fish do not need Visas (16/40 character maximum)"
output:
  word_document:
    # reference_docx: ./Manuscript_Instructions/science_template.docx
  pdf_document:
    fig_caption: yes
geometry: margin=1in
editor_options:
  chunk_output_type: inline
# bibliography: ./References/FishForVisa_Ref.bib
# csl: ./References/science.csl
---

@@

Distributions of marine species around the world are not bounded by human-made boundaries; rather shaped by contemporary biotic and abiotic factors, as well as their evolutionary history [@Hunch; @NelsonFishesOfTheWorld]. However, management of living marine resources is predicated on the definition of “stocks” that are delineated by human-made spatial boundaries that often do not correspond to biologically-meaningful population units [@Song:2017iua]. The delineation of Economic Exclusive Zones (EEZs) under the United Nations Convention on the Law of the Seas (UNCLOS) in the early 80s, for instance, established political boundaries across the distribution of many species, creating shared species between nations. Theories and empirical experience demonstrate fisheries targeting resources that straddle between political boundaries complicate fisheries management and potentially reduce its effectiveness to achieve stated objectives.. For example, fishing quotas set by different nations for a shared fishery resource have led to conflict [@Spijkers:2017ij]. Climate change, which is shifting species distributions, is likely to exacerbate this type of conflict [@Pinsky:2018cb] and presents further challenges for fisheries management [@Mills].

Forty years after the formal adoption of UNCLOS, accurate estimates of the number of exploited marine species that distribute across two or more EEZs (hereafter called shared species) in the world is still unclear. Caddy in 1997 estimated that there could be up to 1,500 transboundary fish species in the world. However, such estimation lacked a proper assessment due to limited information on the distribution of species at the time [@Caddy:1997ue]. More recently, Teh et al. (2010) estimated that 206 species can be considered shared, accounting for a total catch of 34.2 million t and a global landed value of USD 30.7 trillion. While this represents a good starting point, the analysis was based on a literature review of shared species and did not include a mechanistic way of determining the transboundary nature of the species [@Teh:2015gd]. Getting accurate estimates of the number of shared species and where they are located can help develop fisheries management and conservation plans that account for the challenges from transboundary resources management [@Millis].

## Methods summary (250/250 words)

This paper aims to identify the world current transboundary species and determine their contribution to the catch and revenue of the world marine fisheries. The Food and Agricultural Organization (FAO) recognizes three types of shared stocks: (i) transboundary stocks, shared by 2 or more neighboring coastal nations; (ii) straddling stocks, occurring in two or more adjacent national jurisdictions and the high seas; and (iii) highly migratory stocks, found in the EEZs of coastal nations that are not necessarily adjoining, and the high seas. We focus on estimating the number of species that meet the first category. We overlay the known distribution of **968** commercially valuable marine fish species and the Exclusive Economic Zones of **175** coastal countries to determine the total number of transboundary species and their contribution to global catch and landed value. We carry out the analysis at the species level, rather than stocks (populations within a species), due to lack of stock-specific spatial and ecological information on all of the fished species [@Teh:2015gd].

We used four-criteria that combined multiple species distribution data-sources and the world's EEZs to determine if a species can be considered a “shared species”. The first criteria only considered shared species between neighbouring EEZs, rather than the extended distribution. We sub-divided EEZs by regions (e.g. Mexico Pacific and Mexico Atlantic). Current species distributions were predicted from occurrence records obtained from publicly available observational databases of the world’s marine species [@REFNEreus] and two structurally distinct species distribution models [@REFGabs; @Zeller] (see Methods). For the second criteria we considered a species to be present in a grid cell if all data sources predicted positive occurrence. The third criteria determined whether the species were exploited at the grid cell using spatially explicit catch data [@Zeller]. A species was kept if both catch data and predicted species occurrences were positive at the same spatial cells. Finally, the last criteria determined a spatial treshold to confirm that a species occurs in both neighboring EEZs. Once we have identified the shared stocks, we estimated the contribution of the species to fishing revenues and landings using the SAU database [@Zeller] and categorized them according to catch trends [@Kleisner:2013gh].

## Results and discussion (~500/1000 words)

Overall, we estimated **633** transboundary species in the world, more than double the number previously estimated [@Teh:2015gd], accounting for **65.4**% of the species analyzed (Fig. 1A). Between 2005 and 2014, national fleets targeting shared stocks within EEZs landed an average of **69.2** million tonnes, representing **64**% of the EEZ-global catch, and generating USD **79.7** million in fishing revenue, equivalent to **49.8**% of the total revenue of the same period. These values are higher than previously estimated [@Teh] and higlight the importance of these stocks at the global level.

Transboundary species are particularly important at the regional and national level. In terms of catch and revenue, Eastern Asia and Latin America and the Caribbean are the higher zones (Fig 1B). Particular for counties in North America (Bermuda, Canada, Greenland, and the US), a moderate amount of catch (3.4 thousand tons) are worth 11.7 billion USD, highlighting the importance of transboundary species to the economy of the region. The EEZs along Western Eurpe, in turn, are home of **X** species responding to the joint fisheries management policy established by the European Union. In here, quotas are allocated by species and regulatory area [@Melnychuk:2017gh], making this region **unique** in terms of joint management of a high numer of shared stocks.

At the country level, transboundary species are most valuable for the United States (USD 9,605 million), Russia (USD 6,361 million), China (USD 6,279 million) and Peru (USD 5,211). On the other hand, EEZs like that of Portugal, the Mediterranean side of Spain, Senegal, and the UK are rich in transboundary species (n > 110), but these do not represent a large amount of revenue, comparativeley. China, the world's top fish producer with 15.3 million tons landed in 2016 [@SOFIA2018], is the only country categorized high in both number of transboundary species (n = 119) and revenue. For the US, 16 species make up 90% of the revenue from transboudary species, with American scallop (*Placopecten magellanicus*) shared with Canada in Georges Bank [@https://elischolar.library.yale.edu/cgi/viewcontent.cgi?article=1008&context=fes-pubs] contributing 41.8% or the revenue. Russia shares the largest fished species in the world, Alaskan polllock (*Theragra chalcogramma*) [@SOFIA2018], with both Japan and the US. For Russia, this fishery is responsible for 50% of the revenue from transboundary species with other eight species comprising the 90% of revenue. In the case of Peru, Peruvian anchoveta (*Engraulis ringens*) and jumbo squid (*Dosidicus gigas*), both shared with Chile, represent almost 90% of the revenue from transboundary species, with anchoveta accounting for 79%.

These results hilights the importance of transboundary species not only at the global and regional levels, but also at the country level.

@@

We found transboundary fisheries to be of substantial economic importance in regions with past fisheries-related conflicts [@Spijkers:2019cz]. For example, the United States and Canada had a dispute over Pacific salmons (*Oncorhynchus sp*) that lasted over 10 years and was only resolved after these two nations set a series of regulations to compensate losses [@Miller2003]. Since 2007, the European Union, Norway, Iceland, and the Faroe Islands have been at odds over the size and relative allocation of total allowable catches (TACs) for mackerel [@Spijkers:2017ij], a species that is transboundary and contributes **X** to landings and **X** in revenue to the countries involved in the dispute. In addition, climate change could exacerbate conflicts over transboundary species as climate change shifts the distribution of marine species around the globe [@Pinsky].

@@

We determine whether or not a species can be considered transboundary. The method is based on a series of criteria to be met and the overlay of species distributions and the spatial boundaries of the world’s EEZs. For the current analysis we work at the species level and adopt the Food and consider transboundary species as those that occur within the Exclusive Economic Zone (EEZ) of neighboring countries [@Munro:2002uf; @FAO]. All data was scaled to 0.5 degree latitude x 0.5 degree longitude grid cells (*n* = ~180,000 grid cells).

@@

We use the occurrence database, the ENM-Nereus model, and SDM-SAU model to determine the presence of each species within each of the 0.5 x 0.5 grid cells. We computed a *Species Index* by dividing the number of datasets that confirm presence of a species in a grid cell over the total amount of datasets (*n* = 3). The index is unit-less and represents a scale from 0.3 to 1 where 0.3 means only one dataset reports presence of the species in a grid cell, 0.6 represents two of three sources, and 1 means agreement between all datasets. For the analysis, we selected only those cases where *Species Index* = 1 to get a more conservative estimate of transboundary species (**S1_Map_per_dataset**).

@@

# Recicled text (Not to be included)

<!-- From 1973 to 1982, members of the United Nations held a series of meetings to discuss regulations regarding the high seas, a region of international common property, at that time, consisting of waters from 12 miles from shore [@McRaeandMunro1989]. The establishment of the UN Convention on the Law of the Sea (UNCLOS) allowed coastal states to claim jurisdiction over the exploration and exploitation of marine resources over 200 nautical miles off their coasts [@UN:1982]. While intended to improve fisheries management by granting property rights over shared resources, this arbitrary delimitation of management areas was not informed by bio-geography, essentially ignoring species distributions and creating what we know today as shared fish stocks [@Gulland, @Munro2002]. Fish stocks, are comonly refered in fisheries management as a particular population that is more or less isolated from other stocks of the same species and hence self-sustaining [@FAO:1999tk]. The Food and Agricultural Organization (FAO) recognizes three types of shared stocks: (*i*) transboundary stocks, shared by 2 or more neighboring coastal nations; (*ii*) straddling stocks, occurring in two or more adjacent national jurisdictions and the high seas; and (*iii*) highly migratory stocks, found in the EEZs of coastal nations that are not necessarily adjoining, and the high seas. The inception of shared stocks called for the establishment of new fisheries management methods. Managers and scientists adopted a game theory approach – or means to analyze strategic interactions among decision-makers – which demonstrates that collaboration is most likely to result in the best overall outcome for nations sharing a common resource [@Bailey:2010ga,@Eide:2013dh]. Lack of collaboration in shared stocks may threaten stock sustainability [@Clark:1980js; @Nguyen:2018dk], reduce the profitability potential of the fishery [@Merino:2007jz], and lead to conflict between coastal nations [@Spijkers:2017ij], although cooperation and non-cooperation can be more complex [@Jensen:2015cf]. Moreover, under climate change, species distributions will shift [@Cheung2016], resulting in new shared stocks [@Pinsky:2018cb] and highlighting the key role cooperative fisheries management can play in maintaining stocks and profitability [@Nguyen:2018dk] as well as hopefully reducing fisheries conflict [@Spijkers:2019cz]. -->

<!-- There is a gap in assessing the nature and number of transboundary species in the world based on their spatial distribution. Caddy [@Caddy:1997ue] in 1997 estimated that there could be up to 1,500 transboundary fish species in the world. However, such estimation lacked a proper assessment due to limited information on the distribution of species at the time [@Caddy:1997ue]. More recently, Teh et al. [@Teh:2015gd] estimated that 206 species can be considered shared, accounting for a total catch of 34.2 million t and a global landed value of USD 30.7 trillion. While this represents a good starting point, the analysis was based on a literature review of shared species and did not include a mechanistic way of determining the transboundary nature of the species [@Teh:2015gd]. Thus, despite recent research highlighting the interconnection of marine species [@Ramesh:2019va], the total number of transboundary species is yet to be identified at a global scale. Yet, such an understanding is critical to effectively managing fisheries, as needs differ substantially between discreet and shared species [@Mills] and climate change will yield new challenges associated with new shared fisheries [@Pinsky:2018cb]. -->
